Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-21 draft.tex
.

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
\usepackage[a4paper,margin=1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}
\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}

@@

@@

As an additional test, we applied external poisson input for $\mathrm{t\geq 300\, s}$ to two excitatory subgroups to see whether changes in the firing rate of these subgroups is levelled out by homeostasis or retained. Members of these subgroups were picked randomly and independent of their position. As shown in Figure \ref{Ext_Input_Switch}, differences in among subgroups caused by different input signal strengths are gradually eliminated by the homeostatic mechanism.
\begin{figure}
\begin{center}
%% plot_external_input_switch_activity.py
\includegraphics[width=0.7\textwidth]{../../plots/external_input_switch.png}
\end{center}
\caption{Excitatory population firing rate of subgroups subject to external poisson input for $\mathrm{t\geq 300\, s}$. Red and green subgroups consist of 100 randomly picked exc. neurons each. The remaining 300 neurons (blue) received no external input.}
\label{Ext_Input_Switch}
\end{figure}

@@

\subsection{The Firing Rate Distribution}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. Though the distribution did approximately resemble a log-normal-like shape, it did not perfectly match our expectations. In order to acquire a deeper understanding of the mechanisms that determine these statistics, we developed analytical expression that allowed us to reliably predict the steady-state of firing rates based on the spatial structure of excitatory neurons. We would like to emphasize that this explanation differs from common approaches of understanding the emergence of heavy-tailed statistics of firing rates in recurrent networks. Theoretical studies on the behaviour of recurrent networks are largely based on the assumption that the excitability of neurons is not individually fine-tuned but rather randomly distributed and adjusted globally to achieve a desired mean firing rate. Predictions about the resulting activity are then made based on (potentially simplified) statistical features of recurrent connectivity and the specific form of the neuronal transfer function used in the model, see e.g. \cite{Roxin_Firing_Rate_Distribution,Vreeswijk1998}. Our interpretation of our results leads to the conclusion that this "canonical" approach matches the limiting case of instant diffusion: While Figures \ref{Fir_Rate_Dist_Instant_compare} and \ref{Fir_Rate_Dist_Compare} did not reveal many differences in terms of overall statistics, the decorrelation discussed in Section \ref{Section_Rand_Mat_vs_Sim} and plotted in Figure \ref{Corr_Coeff_vs_D} suggests that increasing the diffusion constant goes along with a transition between a spatially determined configuration of excitatory firing rates and a network topology-determined behaviour.

@@

As an indirect consequence of spatial structure determining activity, we found that neurons with a low-density neighbourhood have a higher chance of developing strong outgoing weights. Since this effect is directly linked to the low-density/high-firing rate-relation shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (C), all previous thoughts regarding plausibility similarly apply to this result: \emph{If} a relationship between density and firing rates existed, we could indeed argue to also expect a relation as shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (D), since it is known that neurons above-average firing rates tend to undergo long-term potentiation \cite{Sjoestroem_Syn_Plasticity_2001,Feldman_STDP_2012}.

@@

First, in Section \ref{Syn_Weight_Dist_Section}, we presented our results regarding the distribution of weights, which turned out to be broader but still approximately log-normal distributed for diffusive homeostasis. As a first interpretation, we hypothesized that this might due to a slightly decreased amount of STDP-induced potentiation (Figure \ref{STDP_change}) in the case of diffusive homeostasis in the first half of the simulation. Another explanation one might consider is that the increased firing rate for some neurons in the case of diffusive homeostasis leads to a greater amount of fluctuation because of the increased amount of STDP-events. However, since the overall variance of STDP-induced fluctuations is not significantly higher but even slightly lower compared to non-diffusive homeostasis, this explanation seems rather unlikely.

Taking a closer look at the distributions depicted in Figure \ref{Weight_Dist}, we noted that

>>>>

\documentclass[10pt,a4paper]{article}
%\usepackage{layouts}
\usepackage[a4paper,margin=1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{caption}
\captionsetup[table]{singlelinecheck=false}
\captionsetup[figure]{singlelinecheck=false}
\author{Fabian Schubert}

@@

\section{Introduction}
Among many functional properties of the mammalian central nervous system, it is involved in maintaining many physiological properties in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behaviour have been found up till now, and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though, experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

The notion of synaptic homeostatic plasticity subsumes the observation that neurons being more or less active than a certain set point of mean activity are regulated by means of their incoming synaptic weights. Theoretical and experimental studies have suggested that this type of regulation is in a multiplicative manner, thereby retaining proportions between individual synaptic strengths \cite{Syn_Plast_Abbott}. Furthermore, one can regard this form of plasticity as an antagonist to other aspects of synaptic plasticity that affect synaptic strengths in an activity dependent, hebbian fashion. Forms of regulation belonging to intrinsic homeostasis alter the neurons' capability of spiking upon external input by changing "intrinsic" properties of the nerve cell. More precisely, a neurons' ability to form an action potential can change by modifying the conductance of ion channels in its cell membrane \cite[p. 156]{Theor_Neur_Dayan}. It is important to note that - depending on the specific mechanisms involved - both concepts of neural homeostasis can occur on a very wide range of different time scales, ranging from milliseconds to hours or days.

Theoretical models of homeostasis are generally based on a form of feedback control. As such, it necessarily contains an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This raises two questions: First, what would be an appropriate choice of homeostatic targets? Second, how could a neural system with fixed homeostatic targets encode changes of external stimuli if homeostasis overrides any information about changes in synaptic input?

@@

As an additional test, we applied external poisson input for $\mathrm{t\geq 300\, s}$ to two excitatory subgroups to see whether changes in the firing rate of these subgroups is levelled out by homeostasis or retained. Members of these subgroups were picked randomly and independent of their position. As shown in Figure \ref{Ext_Input_Switch}, differences in among subgroups caused by different input signal strengths are gradually eliminated by the homeostatic mechanism.
\begin{figure}
\begin{center}
%% plot_external_input_switch_activity.py
\includegraphics[width=\textwidth]{../../plots/external_input_switch.png}
\end{center}
\caption{Excitatory population firing rate of subgroups subject to external poisson input for $\mathrm{t\geq 300\, s}$. Red and green subgroups consist of 100 randomly picked exc. neurons each. The remaining 300 neurons (blue) received no external input.}
\label{Ext_Input_Switch}
\end{figure}

@@

\subsection{The Firing Rate Distribution} \label{Fir_Rate_Dist_Discussion_Section}
By implementing diffusive homeostasis into the LIF-SORN model we managed to achieve a broader, skewed firing rate distribution while preserving appreciated features of network topology. Though the distribution did approximately resemble a log-normal-like shape, it did not perfectly match our expectations. In order to acquire a deeper understanding of the mechanisms that determine these statistics, we developed analytical expression that allowed us to reliably predict the steady-state of firing rates based on the spatial structure of excitatory neurons. We would like to emphasize that this explanation differs from common approaches of understanding the emergence of heavy-tailed statistics of firing rates in recurrent networks. Theoretical studies on the behaviour of recurrent networks are largely based on the assumption that the excitability of neurons is not individually fine-tuned but rather randomly distributed and adjusted globally to achieve a desired mean firing rate. Predictions about the resulting activity are then made based on (potentially simplified) statistical features of recurrent connectivity and the specific form of the neuronal transfer function used in the model, see e.g. \cite{Roxin_Firing_Rate_Distribution,Vreeswijk1998}. Our interpretation of our results leads to the conclusion that this "canonical" approach matches the limiting case of instant diffusion: While Figures \ref{Fir_Rate_Dist_Instant_compare} and \ref{Fir_Rate_Dist_Compare} did not reveal many differences in terms of overall statistics, the decorrelation discussed in Section \ref{Section_Rand_Mat_vs_Sim} and plotted in Figure \ref{Corr_Coeff_vs_D} suggests that increasing the diffusion constant goes along with a transition between a spatially determined configuration of excitatory firing rates and a network topology-determined behaviour.

@@

@@

\subsubsection{Distribution of Weights}
In Section \ref{Syn_Weight_Dist_Section}, we presented our results regarding the distribution of weights, which turned out to be broader but still approximately log-normal distributed for diffusive homeostasis. As a first interpretation, we hypothesized that this might due to a slightly decreased amount of STDP-induced potentiation (Figure \ref{STDP_change}) in the case of diffusive homeostasis in the first half of the simulation. Another explanation one might consider is that the increased firing rate for some neurons in the case of diffusive homeostasis leads to a greater amount of fluctuation because of the increased amount of STDP-events. However, since the overall variance of STDP-induced fluctuations is not significantly higher but even slightly lower compared to non-diffusive homeostasis, this explanation seems rather unlikely.

Taking a closer look at the distributions depicted in Figure \ref{Weight_Dist}, we noted that the increased width of the distribution is largely due to to an increment on the left half of the curve. Since new weights are initialized at $\mathrm{10^{-4}\, mV}$, this might indicate that some weights are drifting much more slowly from their value of initialization compared to other weights. These "slow drifters" are presumably connecting neurons both spiking at a below-average firing rate. On the other hand, for non-diffusive homeostasis, one can assume that the amount of STDP-fluctuations is approximately equal for all recurrent excitatory weights because of the fixation of firing rates.

\subsubsection{Strong Differences in Outgoing Weights}
The most prominent significant difference between homeostatic mechanisms we found within network topology was the emergence of significantly above-average mean outgoing weights for a small subset of neurons. By comparing the given mean values to statistics acquired by randomly shuffling recurrent excitatory weights (see Figure \ref{Out_Weight_Mean_Quantile}), we found that this effect cannot be explained by the overall increase of weight heterogeneity that we discussed in the previous section. Thus, we identify the over-representation of strong outgoing weights as a non-random effect. However, we would like to emphasize that diffusive homeostasis is only indirectly responsible in the sense that it allows for a sustained heterogeneity among firing rates.

As an indirect consequence of spatial structure determining activity (as discussed in Section \ref{Fir_Rate_Dist_Discussion_Section}), we found that neurons with a low-density neighbourhood have a higher chance of developing these strong outgoing weights. Since this effect is directly linked to the low-density/high-firing rate-relation shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (C), all previous thoughts regarding plausibility similarly apply to this result: \emph{If} a relationship between density and firing rates existed, we could indeed argue to also expect a relation as shown in Figure \ref{Inverse_Dens_vs_Sum_Out_Weights} (D), since it is known that neurons with above-average firing rates tend to undergo long-term potentiation \cite{Sjoestroem_Syn_Plasticity_2001,Feldman_STDP_2012}.
